Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on reported case numbers

### 2021-04-19 manuscript/v7_submission_revision_2/replyreviewer.Rmd

---
title: "Reply to Reviewer #2 / Journal Requirements"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We want to thank the reviewer for the feedback. We revised the manuscript to address the remark of the reviewer regarding social determinants of COVID-19 infections:

**I find the topic of the paper very important and the methodology very interesting. The author has used a number of variables to analyse possible determinants of COVID-19 infections. While the paper acknowledges that there are social factors which affect the spread of the COVID-19 this is not properly discussed in the paper. I strongly suggest the author to refer to the literature on Social Determinants of Health when discussing social variables. For example see the recent paper by Galanis and Hanieh in Social Science and Medicine on Incorporating Social Determinants of Health into Modelling of COVID-19 and references therein.**

We extended the Discussion section of our manuscript to address this rightful remark. More specifically, we included references to reflect on the situation of social determinants on COVID-19 infections in Germany, most prominently the socio-economic status of households.

Furthermore we reply to the Journal Requirements regarding references within the manuscript:

**Please review your reference list to ensure that it is complete and correct. If you have cited papers that have been retracted, please include the rationale for doing so in the manuscript text, or remove these references and replace them with relevant current references. Any changes to the reference list should be mentioned in the rebuttal letter that accompanies your revised manuscript. If you need to cite a retracted article, indicate the article’s retracted status in the References list and also include a citation and full reference for the retraction notice.**

We checked all references, as of 19 April 2021 none were retracted since the first version of the manuscript. We updated some references that moved from a pre-print status to being published after peer-review. Furthermore, we cleaned up the reference list from faulty ISSN.

### 2021-04-26 manuscript/v7_submission_revision_2/replyreviewer.Rmd

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We want to thank the reviewer for the feedback. We revised the manuscript to address the remark of the reviewer regarding social determinants of COVID-19 infections:

**I find the topic of the paper very important and the methodology very interesting. The author has used a number of variables to analyse possible determinants of COVID-19 infections. While the paper acknowledges that there are social factors which affect the spread of the COVID-19 this is not properly discussed in the paper. I strongly suggest the author to refer to the literature on Social Determinants of Health when discussing social variables. For example see the recent paper by Galanis and Hanieh in Social Science and Medicine on Incorporating Social Determinants of Health into Modelling of COVID-19 and references therein.**

We extended the Discussion section of our manuscript to address this rightful remark. More specifically, we included 3 additional references to reflect on the situation of social determinants on COVID-19 infections especially in Germany, most prominently the socio-economic status of households.

### 2021-04-26 manuscript/v6_medrxiv_revision_2/discussion.Rmd
updates reply reviwer and social determinants

A recent review on COVID-19 based on evidence from the US and UK concludes that low socio-economic status groups are being hit harder by the pandemic [@Wachtler2020Socioeconomic]. Albeit specific pathways remain unclear, many studies found associations with poverty or its correlates such as poor and potentially overcrowded housing conditions. For Germany, a higher case fatality of COVID-19 cases in districts with higher socio-economic deprivation has also been reported just recently, which was especially pronounced in the second wave of the pandemic [@rki2021sozialeunterschiede]. Similarly, our analysis identified a decreasing effect on COVID-19 case numbers within districts with a higher socio-economic status during the first wave.

@@

>>>>

A recent review on COVID-19 based on evidence from the US and UK concludes that low socio-economic status groups are being hit harder by the pandemic [@Wachtler2020Socioeconomic]. Albeit specific pathways remain unclear, many studies found associations with poverty or its correlates such as poor and potentially overcrowded housing conditions. For Germany, a higher case fatality of COVID-19 cases in districts with higher socio-economic deprivation has also been reported just recently, which was especially pronounced in the second wave of the pandemic [@DAE218459]. Similarly, our analysis identified a decreasing effect on COVID-19 case numbers within districts with a higher socio-economic status during the first wave.

@@

Social determinants of health are important factors to consider in an epidemiological framework of a pandemic disease like COVID-19. To account for this problem, we included several socio-economic confounders that were available on a district level in Germany. While our analysis is not an exhaustive analysis of the effects of social determinants on COVID-19 infections, we emphasize the necessity of their inclusion and our results add to the growing body of evidence that these factors interact with each other and cluster especially among people or within areas of underprivileged conditions, with detrimental effects on population health [@solar2010conceptual].

### 2021-04-26 manuscript/v6_medrxiv_revision_2/manuscript.tex
updates reply reviwer and social determinants

@@

>>>>

A recent review on COVID-19 based on evidence from the US and UK concludes that low socio-economic status groups are being hit harder by the pandemic \citep{Wachtler2020Socioeconomic}. Albeit specific pathways remain unclear, many studies found associations with poverty or its correlates such as poor and potentially overcrowded housing conditions. For Germany, a higher case fatality of COVID-19 cases in districts with higher socio-economic deprivation has also been reported just recently, which was especially pronounced in the second wave of the pandemic \citep{DAE218459}. Similarly, our analysis identified a decreasing effect on COVID-19 case numbers within districts with a higher socio-economic status during the first wave.

@@

Social determinants of health are important factors to consider in an epidemiological framework of a pandemic disease like COVID-19. To account for this problem, we included several socio-economic confounders that were available on a district level in Germany. While our analysis is not an exhaustive analysis of the effects of social determinants on COVID-19 infections, we emphasize the necessity of their inclusion and our results add to the growing body of evidence that these factors interact with each other and cluster especially among people or within areas of underprivileged conditions, with detrimental effects on population health \citep{solar2010conceptual}.
